Sampbias, a method to evaluate geographic sampling bias in species distribution data

### 2019-12-03 manuscript.rmd
Update manuscript.rmd

# Abstract
Georeferenced species occurrences from public databases have become essential to biodiversity resaerch and conservation, but have limitations. Geographically biased sampling is a widely recognized issue that might severely affect analyses. Especially "roadside bias", i.e. differences in sampling intensity among localities caused by differences in accessibility for humans is ubiquitous and might differ in strength among taxonomic groups and datasets. Yet, no general methodology existed to quantify the effect of roadside bias on a dataset level. Here we present *sampbias*, a novel algorithm and software to estimate the biasing effect of accessibility (by roads, rivers, airports, cities, or any user-defined structures) in species occurrence datasets. *Sampbias* is based on a null model of equal sampling and exponential distance decay and the resultsare compareable among biasing factors and datasets. *Sampbias* is implemented as a user-friendly R package, and Shiny app. We exemplify the use of *sampbias* on a dataset of  mammal occurrences from the Indonesian island of Borneo, downloaded from www.gbif.org. *Sampbias* offers an efficient and largely automated means for biodiversity scientists and non-specialists alike to explore bias in species occurrence data. The output of *sampbias* may be used to identify priorities for further collection or digitalization efforts, provide bias surfaces for species distribution modelling, or assess the reliability of scientific results based on publicly available species distribution data.

@@

# Introduction
Publicly available datasets of geo-referenced species occurrences, such as provided by the Global Biodiversity Information Facility (www.gbif.org) have become a fundamental resource  in biological sciences, especially in biogeography, conservation, macroecology, and systematics. However, because these dataset are "presence-only" data, rarely include information on collection effort, are not collected systematically and often compiled from a variety of collection purposes (e.g. scientific expeditions, census counts, genetic barcoding studies, and citizen science observations) they are subject to collection biases [@Meyer2016]. That is the number of data points available is biased by factors other than species' presence or abundance, including the under-sampling of specific taxa ("taxonomic bias", e.g., birds *vs.* nematodes), specific geographic regions ("geographic bias", i.e. easily accessible *vs.* remote areas), and specific temporal periods ["temporal bias", i.e. wet season *vs.* dry season, @Isaac2015; @Boakes2010]. While these biases are broadly recognized, and approaches exist to account for them in some analyses (for instance for species-richness estimation [@Engemann2015] species distribution modelling [@Stolar2015;@Beck2014; @Fithian2014; @Warren2014; @Boria2014; @Varela2014; @Fourcade2014], occupancy modelling [@Kery2016], or abundance estimations [@Shimadzu2015]), few attempts have been made [but see @Ruete2015] to discern among different sources of bias or to compare the strength of bias among datasets.

@@

The implications of not considering spatial collection bias in biodiversity research are likely to be substantial [@Meyer2016; @Rocchini2015; @Shimadzu2015;  @Yang2013; @Kramer-Schadt2013; @Barbosa2013]. While it is unrealistic to expect that spatial biases in biodiversity data will ever disappear, it is crucial that researchers realise the intrinsic biases associated with the biodiversity data they are dealing with. This is the first step towards estimating to which extent these biases may affect their analyses, results and conclusions drawn from such data. Therefore, it is advisable for any study dealing with species occurrence data to assess the strength of accessibility bias in the underlaying data.

@@

# Description
## General concept
Under the assumption that organisms exist across the entire area of interest, we can expect the number of sampled occurrences to be distributed uniformly in space (even though, of course, the density of individuals and the species composition may be heterogeneous).  With *sampbias* we assess if a set of occurrences significantly departs from a null uniform distribution and whether these discrepancies between expected and observed distributions can be explained by distance from factors that potentially bias their sampling probability (e.g. distance from cities or roads).

@@

## Distance calculation
*Sampbias* uses gazetteers of the geographic location of bias sources (e.g. roads) to generate a grid across the study area (the geographic extent of the dataset) for each gazetteer and then calculates the distance ("as the crow flies") of the midpoint of each grid cell to the closest cell containing an instance of the gazetteer. We then use these distance grids to sample the distribution of distances in the observed dataset and the null distribution in a reference dataset of equal size with randomly distributed records (the null model). The resolution of the grid defined the precision of the distance estimates, for instance a 1x1 degree raster will yield approximately a 100km precision at the equator.

## Quantifying accessibility bias using maximum likelihood
Given the placement of a particular bias sources in the area of interest and assuming a uniform distribution of samples, the probability of a sampled occurrence located at a distance $d$ from the closest bias source is a function of the amount of available area at that distance. That is, the larger the area located at distance $d$ from a bias source the more samples we expect. For simplicity we discretize the area of interest in a number of grid cells and indicate with $f(x)$ the function describing the number of available grid cells at any distance $x$, for $0 < x < max(x)$, where $max(x)$ is the maximum observed distance from a bias source. The function $f(x)$ is therefore calculated based on the distances of each grid cell from its closest bias source.

The distribution of samples, in the absence of bias, should therefore represent a random sample from $f(x)$ and reflect its shape, i.e. $d ~ f(x)$ (Fig. \ref{fig:distdi}). However, in the presence of a bias, we expect the probability of finding occurrences to decrease with increasing distance from a bias source. This in turn will alter the resulting distribution of samples that no longer match the expected distribution. Here, we model the effect of sampling bias by assuming that the probability of sampling an occurrence decreases exponentially with increasing distance (Fig. \ref{fig:distdi}), following the function $b(x,l) = l exp(-lx)$, where $l$ is the rate parameter. Under these assumption the expected distribution of samples is given by $g(x,l) \propto f(x) b(x,l)$.

The rate parameter $l$ describes the strength of the bias effect. When  $l$ is large the expected probability of sampling occurrences decreases very quickly as you move away from a bias source (Fig. \ref{fig:distdi}). In contrast, when the $l$ parameter is small, the resulting exponential distribution effectively becomes more and more similar to a uniform distribution (Fig. \ref{fig:distdi}), indicating that increasing distance from e.g. a city does not affect the sampling probability. We treat $l$ as an unknown variable and estimate it using maximum likelihood from the data based on the probability density function described by $g(x,l)$. This essentially means finding the value of $l$ that best explains any discrepancies between the expected distribution ($f(x)$) and the observed occurrences.

Once we have an estimated value of $l$, we can infer the expected accessibility bias as a function of distance using $b(x,l)$. Since the function $b(x,l)$ describes the exponentially decreasing sampling probability in relation to distance, we can define a standardized bias function as: $B(x,l) = 1 - b(x,l)/b(0,l)$, where  the level of bias is set to 0 at distance 0 from the bias source. The standardized bias function tends asymptotically to 1 as the distance tends to infinity. However, since in any area $max(x) << \infty$, for small values of $l$ (i.e. little or no bias), $B(x,l)$ will look essentially like a uniform distribution with values very close to 0. Large values of $l$ (i.e. strong bias) will instead result in a curve that quickly approaches 1 with increasing distance. The values provided by the standardized bias function can be interpreted as the proportion of occurrences that are missing from the sample, compared to the observed samples at distance 0. Thus, if for a given estimated $l$ we have $B(50,l) = 0.20$, we can expect that at 50 Km from e.g. a road the number of occurrences per grid cell will be about 80% of the occurrences sampled at distance 0 from the road, with 20% missing due to sampling bias.

>>>>

# Abstract
Georeferenced species occurrences from public databases have become essential to biodiversity research and conservation, but have limitations. Geographically biased sampling is a widely recognized issue that might severely affect analyses. Especially "roadside bias", i.e. differences in sampling intensity among localities caused by differences in accessibility for humans is ubiquitous and might differ in strength among taxonomic groups and datasets. Yet, no general methodology exists to quantify the effect of roadside or other sources of bias on a dataset level. Here we present *sampbias*, a novel algorithm and software to estimate the biasing effect of accessibility (by roads, rivers, airports, cities, or any user-defined structures) in species occurrence datasets. *Sampbias* is based on a null model of even sampling and assesses whether instead sampling probability decays exponentially with distance. The results are comparable among biasing factors and datasets. *Sampbias* is implemented as a user-friendly R package, and Shiny app. We exemplify the use of *sampbias* on a dataset of  mammal occurrences from the Indonesian island of Borneo, downloaded from www.gbif.org. *Sampbias* offers an efficient and largely automated means for biodiversity scientists and non-specialists alike to explore bias in species occurrence data. The output of *sampbias* may be used to identify priorities for further collection or digitalization efforts, provide bias surfaces for species distribution modelling, or assess the reliability of scientific results based on publicly available species distribution data.

@@

# Introduction
Publicly available datasets of geo-referenced species occurrences, such as provided by the Global Biodiversity Information Facility (www.gbif.org) have become a fundamental resource  in biological sciences, especially in biogeography, conservation, macroecology, and systematics. However, because these datasets are "presence-only" data, they rarely include information on collection effort. Instead they are typically not collected systematically and often compiled from a variety of sources (e.g. scientific expeditions, census counts, genetic barcoding studies, and citizen-science observations),  thus becoming subject to collection biases [@Meyer2016]. 
That is the number of data points available is biased by factors other than species' presence or abundance, including the under-sampling of specific taxa ("taxonomic bias", e.g., birds *vs.* nematodes), specific geographic regions ("geographic bias", i.e. easily accessible *vs.* remote areas), and specific temporal periods ["temporal bias", i.e. wet season *vs.* dry season, @Isaac2015; @Boakes2010]. While these biases are broadly recognized, and approaches exist to account for them in some analyses (for instance for species-richness estimation [@Engemann2015] species distribution modelling [@Stolar2015;@Beck2014; @Fithian2014; @Warren2014; @Boria2014; @Varela2014; @Fourcade2014], occupancy modelling [@Kery2016], or abundance estimations [@Shimadzu2015]), few attempts have been made to discern among different sources of bias or to compare the strength of bias among datasets [but see @Ruete2015].

@@

The implications of not considering spatial collection bias in biodiversity research are likely to be substantial [@Meyer2016; @Rocchini2015; @Shimadzu2015;  @Yang2013; @Kramer-Schadt2013; @Barbosa2013]. While it is unrealistic to expect that spatial biases in biodiversity data will ever disappear, it is crucial that researchers realise the intrinsic biases associated with the biodiversity data they are dealing with. This is the first step towards estimating to which extent these biases may affect their analyses, results, and conclusions drawn from such data. Therefore, it is advisable for any study dealing with species occurrence data to assess the strength of accessibility bias in the underlaying data.

@@

# Description
## General concept
Under the assumption that organisms exist across the entire area of interest, we can expect the number of sampled occurrences to be distributed uniformly in space (even though, of course, the density of individuals and the species composition may be heterogeneous). 
[[*I think we should acknowledge here that this assumption is valid when looking at a geographically restricted area, eg within a tropical forest. Of course different biomes eg forest, alpine, oceanic will result in different carrying capacity*]] 
With *sampbias* we assess if a set of occurrences significantly departs from a null uniform distribution and whether these discrepancies between expected and observed distributions can be explained by distance from factors that potentially bias their sampling probability (e.g. distance from cities or roads).

@@

## Distance calculation
*Sampbias* uses gazetteers of the geographic location of bias sources (e.g. roads) to generate a grid across the study area (the geographic extent of the dataset) for each gazetteer and then calculates the distance ("as the crow flies") 
[[*meaning?*]]
of the midpoint of each grid cell to the closest cell containing an instance of the gazetteer. We then use these distance grids to sample the distribution of distances in the observed dataset and the null distribution in a reference dataset of equal size with randomly distributed records (the null model). The resolution of the grid defined the precision of the distance estimates, for instance a 1x1 degree raster will yield approximately a 100km precision at the equator.

## Quantifying accessibility bias using maximum likelihood
Given the placement of a particular bias sources in the area of interest and assuming a uniform distribution of samples, the probability of a sampled occurrence located at a distance $d$ from the closest bias source is a function of the amount of available area at that distance. That is, the larger the area located at distance $d$ from a bias source the more samples we expect. For simplicity we discretize the area of interest in a number of grid cells and indicate with $f(x)$ the function describing the number of available grid cells at any distance $x$, for $0 < x < \max(x)$, where $\max(x)$ is the maximum observed distance between a cell and the clostest bias source. The function $f(x)$ is therefore calculated based on the distances of each grid cell from its closest bias source.

The distribution of samples, in the absence of bias, should therefore represent a random sample from $f(x)$ and reflect its shape, i.e. $d ~ f(x)$ (Fig. \ref{fig:distdi}). However, in the presence of a bias, we expect the probability of finding occurrences to decrease with increasing distance from a bias source. This in turn will alter the resulting distribution of samples that no longer match the expected distribution. Here, we model the effect of sampling bias by assuming that the probability of sampling an occurrence decreases exponentially with increasing distance (Fig. \ref{fig:distdi}), following the function $b(x,l) = l \exp(-lx)$, where $l$ is the rate parameter. Under these assumption the expected distribution of samples is given by $g(x,l) \propto f(x) b(x,l)$.

[[*I wonder if we shouldn't use " $b(x,l) = \exp(-lx)$" so the Y-value is always 1 at a distance of 0. That would require normalizing the likelihood based on the integral of the curve though, as it will be different from 1* ]

The rate parameter $l$ describes the strength of the bias effect. When  $l$ is large the expected probability of sampling occurrences decreases very quickly as you move away from a bias source (Fig. \ref{fig:distdi}). In contrast, when the $l$ parameter is small, the resulting exponential distribution effectively becomes more and more similar to a uniform distribution (Fig. \ref{fig:distdi}), indicating that increasing distance from e.g. a city does not affect the sampling probability. We treat $l$ as an unknown variable and estimate it using maximum likelihood from the data based on the probability density function described by $g(x,l)$. This essentially means finding the value of $l$ that best explains any discrepancies between the expected distribution ($f(x)$) and the observed occurrences.
[[*I wonder if this isn't in fact a posterior probability where f(x) is the prior and b(x,l) the likelihood or something like that*]]

Once we have an estimated value of $l$, we can infer the expected accessibility bias as a function of distance using $b(x,l)$. Since the function $b(x,l)$ describes the exponentially decreasing sampling probability in relation to distance, we can define a standardized bias function as: $B(x,l) = 1 - b(x,l)/b(0,l)$, 
[[*Actually I think this is effectively equivalent to what I worteabove $b(x,l) = \exp(-lx)$*]]
where  the level of bias is set to 0 at distance 0 from the bias source. The standardized bias function tends asymptotically to 1 as the distance tends to infinity. However, since in any area $max(x) << \infty$, for small values of $l$ (i.e. little or no bias), $B(x,l)$ will look essentially like a uniform distribution with values very close to 0. Large values of $l$ (i.e. strong bias) will instead result in a curve that quickly approaches 1 with increasing distance. The values provided by the standardized bias function can be interpreted as the proportion of occurrences that are missing from the sample, compared to the observed samples at distance 0. Thus, if for a given estimated $l$ we have $B(50,l) = 0.20$, we can expect that at 50 Km from e.g. a road the number of occurrences per grid cell will be about 80% of the occurrences sampled at distance 0 from the road, with 20% missing due to sampling bias.
